Synthesising Concise Termination and Non-Termination Proofs for Bit-Vector Programs

### 2014-07-08 papers/termination/paper.tex
Move CEGIS algorithm
Fix bad hboxes

We can prove that the program is unconditionally terminating by
finding a ranking function for its transition relation.
%
\begin{definition}[Ranking function]
%
An injective function ${R:X\to Y}$ is a \emph{ranking function} for the
transition relation $T$ if $Y$ is a well-founded set with order $>$ and 
$R$ is monotonically decreasing with respect to $T$.  That is
to say:
$$\forall x, x' \in X. T(x, x') \Rightarrow R(x) > R(x')$$
%
\end{definition}

@@

In the context of termination analysis, a witness to the
satisfiability of a generalised termination specification is a
\mbox{(non-)termination} proof.  As mentioned earlier in the paper, our
analysis must ensure precise treatment of bit-wise operations and arithmetic
modulo fixed widths by modelling bit-accurate semantics.  The synthesiser
must therefore be able to compute solutions over machine
integers and floating-point numbers from the specifications detailed in
Section~\ref{sec:second.order}.  In the next section, we will begin our treatment of
bit-accurate synthesis by illustrating some of the differences between
machine arithmetic and conventional integer arithmetic.

@@

%\begin{figure}
\begin{algorithm}
{\small
 \caption{Abstract refinement algorithm
 \label{alg:cegis}}
 %\vspace{-1.7em}
 %\begin{multicols}{2}
 \begin{algorithmic}[1]
 \Statex
\Function{synth}{{\sc inputs}, $\sigma$}
  \Let{$(i_1, \ldots, i_N)$}{{\sc inputs}}
  \Let{query}{$\exists P, x_0 . \forall i \in \text{\sc inputs} . \sigma(P, x_0, i)$}
  \Let{result}{decide(query)}
  \If{result.satisfiable}
    \State \Return{result.model}
  \Else
    \State \Return{unsatisfiable}
  \EndIf
\EndFunction
\Statex
\Function{verif}{$P, x_0, \sigma$}
  \Let{query}{$\exists x . \lnot \sigma(P, x_0, x)$}
  \Let{result}{decide(query)}
  \If{result.satisfiable}
    \State \Return{result.model}
  \Else
    \State \Return{valid}
  \EndIf
\EndFunction
\columnbreak
\Statex
\Function{Synthesise}{$\sigma$}
  \Let{{\sc inputs}}{$\emptyset$}
  \Loop
    \Let{candidate}{\Call{synth}{{\sc inputs}}}
    \If{candidate = UNSAT}
      \State \Return{unsatisfiable}
    \EndIf
    \Let{res}{\Call{verif}{candidate}}
    \If{res = valid}
      \State \Return{candidate}
    \Else
      \Let{{\sc inputs}}{{\sc inputs} $\cup$ res}
    \EndIf
  \EndLoop
\EndFunction
 \end{algorithmic}
 %\end{multicols}
 }
\end{algorithm}
%\end{figure}

@@

\setlength{\tabcolsep}{16pt}
Integer arithmetic instructions:

@@

\begin{tabular}{lll}
 \verb|le  a b| & \verb|lt  a b| & \verb|sle  a b| \\
 \verb|slt a b| & \verb|eq  a b| & \verb|neq  a b|
\end{tabular}

@@

\begin{tabular}{lll}
 \verb|implies a b| & \verb|ite a b c| & 
\end{tabular}

@@

We use Counterexample Guided Inductive Synthesis
(CEGIS)~\cite{lezama-thesis,sketch, DBLP:conf/iclp/BrainCVF06} to find an
$\mathcal{L}$-program satisfying our specification.  The core of the CEGIS
algorithm is the refinement loop shown in Fig.~\ref{fig:abstract-refinement}
and detailed in Algorithm~\ref{alg:cegis}.  The algorithm is divided into
two procedures: {\sc synth} (see Figure~\ref{fig:synth-dfd}) and {\sc
verif}, which interact via a finite set of test vectors {\sc inputs}.

>>>>

We can prove that the program is unconditionally terminating by
finding a ranking function for its transition relation.
%
\begin{definition}[Ranking function]
%
A function ${R:X\to Y}$ is a \emph{ranking function} for the
transition relation $T$ if $Y$ is a well-founded set with order $>$ and 
$R$ is injective and monotonically decreasing with respect to $T$.  That is
to say:
$$\forall x, x' \in X. T(x, x') \Rightarrow R(x) > R(x')$$
%
\end{definition}

@@

In the context of termination analysis, a witness to the
satisfiability of a generalised termination specification is a
proof of \mbox{(non-)termination}.  As mentioned earlier in the paper, our
analysis must ensure precise treatment of bit-wise operations and arithmetic
modulo fixed widths by modelling bit-accurate semantics.  The synthesiser
must therefore be able to compute solutions over machine
integers and floating-point numbers from the specifications detailed in
Section~\ref{sec:second.order}.  In the next section, we will begin our treatment of
bit-accurate synthesis by illustrating some of the differences between
machine arithmetic and conventional integer arithmetic.

@@

@@

\setlength{\tabcolsep}{14pt}
Integer arithmetic instructions:

@@

\begin{tabular}{lll}
 \verb|le  a b| & \verb|lt  a b| & \verb|sle  a b| \\
 \verb|slt a b| & \verb|eq  a b| & \verb|neq  a b| \\
\end{tabular}

@@

\begin{tabular}{lll}
 \verb|implies a b| & \verb|ite a b c| &  \\
\end{tabular}

\setlength{\tabcolsep}{12pt}

@@

We use Counterexample Guided Inductive Synthesis
(CEGIS)~\cite{lezama-thesis,sketch, DBLP:conf/iclp/BrainCVF06} to find an
$\mathcal{L}$-program satisfying our specification.  The core of the CEGIS
algorithm is the refinement loop shown in Fig.~\ref{fig:abstract-refinement}
and detailed in Algorithm~\ref{alg:cegis}.

%\begin{figure}
\begin{algorithm}
{\small
 \caption{Abstract synthesis algorithm
 \label{alg:cegis}}
 %\vspace{-1.7em}
 %\begin{multicols}{2}
 \begin{algorithmic}[1]
 \Statex
\Function{synth}{{\sc inputs}, $\sigma$}
  \Let{$(i_1, \ldots, i_N)$}{{\sc inputs}}
  \Let{query}{$\exists P, x_0 . \forall i \in \text{\sc inputs} . \sigma(P, x_0, i)$}
  \Let{result}{decide(query)}
  \If{result.satisfiable}
    \State \Return{result.model}
  \Else
    \State \Return{unsatisfiable}
  \EndIf
\EndFunction
\Statex
\Function{verif}{$P, x_0, \sigma$}
  \Let{query}{$\exists x . \lnot \sigma(P, x_0, x)$}
  \Let{result}{decide(query)}
  \If{result.satisfiable}
    \State \Return{result.model}
  \Else
    \State \Return{valid}
  \EndIf
\EndFunction
\columnbreak
\Statex
\Function{Synthesise}{$\sigma$}
  \Let{{\sc inputs}}{$\emptyset$}
  \Loop
    \Let{candidate}{\Call{synth}{{\sc inputs}}}
    \If{candidate = UNSAT}
      \State \Return{unsatisfiable}
    \EndIf
    \Let{res}{\Call{verif}{candidate}}
    \If{res = valid}
      \State \Return{candidate}
    \Else
      \Let{{\sc inputs}}{{\sc inputs} $\cup$ res}
    \EndIf
  \EndLoop
\EndFunction
 \end{algorithmic}
 %\end{multicols}
 }
\end{algorithm}
%\end{figure}

The algorithm is divided into two procedures: {\sc synth} (see Figure~\ref{fig:synth-dfd}) and {\sc
verif}, which interact via a finite set of test vectors {\sc inputs}.
